\LARGE \bf A Hybrid Control Design for Autonomous Vehicles at Uncontrolled Intersections

### 2019-01-30 conference/iv2019/manuscript.tex
ready to submit!

As autonomous vehicles (AVs) inch closer to reality, a central requirement for acceptance will be earning the trust of humans in everyday driving situations. In particular, the interaction between AVs and pedestrians is of
high importance, as every human is a pedestrian at some point of the day. This paper considers the interaction of a pedestrian and an autonomous vehicle at a mid-block, uncontrolled intersection where there is ambiguity
over when the pedestrian should cross and when and how the vehicle should yield. By modeling pedestrian behavior through the concept of gap acceptance, the authors show that a hybrid controller with just four distinct modes allows an autonomous vehicle to successfully interact with a pedestrian across a continuous spectrum of possible crosswalk entry behaviors. The controller is validated through extensive simulation and compared to an alternate POMDP solution and experimental results are provided on a research vehicle for a virtual pedestrian.

@@

While autonomous vehicles have the potential to save thousands of lives every year and create significant societal benefits \cite{Fagnant2015}, widespread adoption is unlikely until AVs gain the broad trust of society. Given that every human is a pedestrian at some point during the day, one of the central ways that autonomous vehicles will be evaluated is through their interactions with pedestrians. In fact, recent events with industry leader Waymo highlight that AV-pedestrian interaction has still not been perfected \cite{MattRichtelandConorDougherty2015}. In general, interactions with pedestrians are complex, even for experienced human drivers. From 2015-2016, pedestrian fatalaties increased by 9\% to 5987, representing the highest number since 1990, and also representing 16\% of all automotive fatalities \cite{HighwayTrafficSafetyAdministration2016}. As autonomous vehicles inch closer to widespread adoption, they must have a clear control strategy for pedestrian interaction that can handle a wide variety of pedestrian behaviors while maintaining a reasonable flow of traffic.

@@

The need to further understand pedestrian behavior for autonomous driving has created a signficant body of research focused on modeling pedestrian behavior given various sensor inputs. Keller et al \cite{Keller2014} presented a study on pedestrian path prediction and action classification (e.g. crossing vs. waiting) using Gaussian process dynamical models and trajectory matching from optical data. Several other techniques for pedestrian trajectory prediction have been proposed, including LQR \cite{Batkovic}, set-based reachability analysis \cite{Koschi2018}, and Markov processes \cite{Karasev2016} Another branch of literature focuses on modeling pedestrian crossing behavior in the form of  video analysis. Schroeder and Rouphail \cite{Schroeder2011} explored factors associated with driver yielding behavior at unsignalized pedestrian crossings. Using logistic regression, the authors found that drivers are more likely to yield to assertive pedestrians who walk briskly in their approach to a crosswalk. Kadali and Perumal \cite{RaghuramKadali2012} studied the ``gap acceptance" behavior of pedestrians at mid-block crosswalks, and found that the gap accepted for crossing was explained by factors such as crossing direction, vehicle speed, and pedestrian age. Yannis et al. \cite{Yannis2013} and Sun et al.\cite{Sun2002} also found that gap acceptance was influenced by the size of the oncoming vehicle and the presence of other pedestrians.

@@

This paper aims to address the above issues by developing a hybrid control architecture that accounts for several distinct pedestrian modes of behavior at an unsignalized crosswalk. An unsignalized crosswalk is chosen as it is the trickiest crosswalk for a pedestrian and vehicle to navigate, as there is no explicit declaration of whose turn it is to cross. Simulations show that the proposed hybrid controller is able to handle a continuous spectrum of pedestrian \textit{gap acceptance} behavior, tolerating a range of highly conservative to highly aggressive pedestrians. Additionally, this paper provides a simulated comparison between the hybrid controller and the solution method proposed by Thornton \cite{Thornton2018}, citing advantages and disadvantages of each method. Finally, experimental results are shown on a real vehicle to demonstrate the feasibility of the proposed controller in real-time.

@@

\begin{figure}
\centering
\includegraphics[width=3.5in]{figures/example2.eps}
\caption{Top: Description of relevant problem variables. Bottom: Schematic of pedestrian approaching a stream of traffic at a mid-block intersection. Ego vehicle shown in green. We develop a hybrid control framework for the green vehicle, which considers the distance ($d$) to a stopping point $\Delta$ ahead of the crosswalk and the pedestrian's position in the crosswalk ($x_p$).}
\label{fig:schematic}
\end{figure}

@@

A diagram of the relevant state variables is shown in Fig.~\ref{fig:schematic}. We assume that knowledge of distance $d(t)$ to a fixed stopping point 3-5 meters ahead of the crosswalk can be estimated by a perception system, and that the vehicle velocity $v(t)$ is known as well. Furthermore, we also assume the perception system is able to determine the position of the pedestrian $x_p$ within the crosswalk, and is able to determine a velocity estimate $\dot{x}_p$ for the pedestrian.

@@

\begin{equation}
a = f(d, v, x_p, \dot{x_p})
\end{equation}

@@

In general, this control problem is difficult as there are multiple stakeholders and therefore multiple competing objectives that must be considered. For example, relevant stakeholders for this problem are the autonomous vehicle occupants, the pedestrian(s), and the other vehicles in the traffic stream.

@@

Similar to time to collision (TTC), the safety gap is a measure of how much time there is before the vehicle would enter the crosswalk if it kept its current speed constant. When faced with a stream of traffic at an unsignalized crosswalk, a pedestrian inherently decides how much of a time gap to accept before crossing. The average gap acceptance is reported in the literature to be between 3-7 seconds, meaning pedestrians usually do not cross if the vehicle would enter the crosswalk in under three seconds \cite{DiPietroCharlesMandKing1970}, and are very likely to cross when they have more than seven seconds \cite{Schmidt2009}.

@@

The proposed control architecture is shown in Fig.~\ref{fig:hybridController}, and the algorithm for operation is shown in Algorithm \ref{alg:hybridController}. At heart, the control algorithm uses a feedback-feedforward methodology to compute desired acceleration commands, but computes these commands differently depending on which of four discrete modes the controller is in. The next section will describe the four modes.

@@

\begin{algorithm} % enter the algorithm environment
\caption{Calculate $a = f(d, v, x_p, \dot{x}_p)$} % give the algorithm a caption
\label{alg:hybridController} % and a label for \ref{} commands later in the document
\begin{algorithmic}
\State $state \leftarrow DRIVING$
\While {$\mathbf{true}$}
\If {$state = DRIVING$}
	\State $a \gets k_s(v-v_\mathrm{speed limit})$ 
	\If{$d > 0 \hspace{1mm} \mathbf{and} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$}
		\If{$\mathrm{timeAdvantage}(x_p, \dot{x_p}, d, v) > t_{max}$} 
		\State $state \leftarrow DRIVING$

@@

\subsection{Nominal Driving}
In the nominal state, the vehicle attempts to drive through the crosswalk at the speed limit $v_\mathrm{speedlimit}$. A simple proportional speed control is applied to keep the vehicle at the speed limit, achieving design constraint 2.a:

@@

The controller remains in the driving mode unless the pedestrian begins to enter the crosswalk before the vehicle has passed. In this case, we mathematically define ``beginning to enter" as the instant the pedestrian begins moving towards the crosswalk, even if the pedestrian is currently on the sidewalk:

@@

 Where $x_F$ is the end of the crosswalk area of interest. Note that according to design constraint 1.b, the definition of $x_F$ will vary according to which US state the vehicle is in. For example, in California, $x_F$ would be at the end of the crosswalk, while in Louisiana, $x_F$ would just be the same half of the crosswalk as the vehicle \cite{Legislatures2018}.

Once the pedestrian enters the crosswalk, the algorithm decides which mode to enter next.

@@

Frequently the vehicle will pass through the intersection well before the pedestrian will reach the vehicle's lane. Assuming the vehicle is operating in a US state where stopping is not explicitly required, design constraint 2.a suggests it is desirable for the vehicle to continue through the intersection.

This can be formalized by calculating the \textit{time advantage} $t_\mathrm{adv}$, also known as the \textit{time to collision} as follows:

@@

Where $x_v$ is the $x$ position of the vehicle in the crosswalk. Algorithm 1 explicitly allows the vehicle to continue driving through the intersection as long as $t_\mathrm{adv}$ exceeds a specified threshold $t_\mathrm{max}$.

@@

In the unlikely event a pedestrian begins to cross when the time gap is low (e.g. under 3 seconds), the vehicle will need to prioritize design constraint 1.a and come to a stop at a deceleration higher than $-a_\mathrm{cmf}$. This occurs when the following conditions holds:

@@

As a final condition, consider the case where the pedestrian enters the crosswalk just as the vehicle is crossing $d = 0$. In this case, it makes little sense to decelerate, as the vehicle has insufficient space to stop before the crosswalk and will risk being rear-ended or stopping in the crosswalk. It makes more sense in this condition for the vehicle to speed up and exit the crosswalk quickly. As a result, if $d < \frac{v}{2a_\mathrm{max}}$, the vehicle speeds up.

@@

Where $d_o$ and $v_o$ are the values of $d$ and $v$ when the controller first enters the yielding state. Inspection shows that the desired velocity is $v_o$ when $d = d_o$ and is 0 when $d = 0$ (recall that we enter the yielding state when $d_o = \frac{\dot{d_o}^2}{2a_\mathrm{cmf}}$).

The controller exits the yielding mode and returns to the driving mode once the pedestrian clears the crosswalk (e.g. once $x > x_F$ in (\ref{eq:crossing})).

@@

In order to check performance of the algorithm, both simulation and experimental studies were undertaken. The simulation results focused on testing Algorithm 1 against the specified design requirements over many pedestrian crossings.

The algorithm is validated in simulation via a built-from-scratch crosswalk simulation environment (Fig. \ref{fig:simFramework}) developed in Python. The simulation environment is built on the ROS robotics interface to enable easy porting of the control architecture to the experimental vehicle in Section \ref{sec:expres}, and for ease of comparison to the method provided by Thornton \cite{Thornton2018}.

\begin{figure}
\centering
\includegraphics[width=2.5in]{figures/simFramework.png}
\caption{Crosswalk simulation environment for algorithm validation. Ego vehicle is shown in black, leading vehicle in traffic stream shown in blue, and pedestrian shown in red.}
\label{fig:simFramework}
\end{figure}

To simulate random pedestrian crossing events, a pedestrian is spawned  on the sidewalk near the entrance to the crosswalk. The pedestrian accepts a randomly sized time gap in the traffic stream to cross the road at constant speed $\dot{x}_p$. While pedestrian speeds vary both according to demographic factors and within an individual crossing attempt \cite{Feliciani2017}, the primary focus of this study is varying the gap acceptance, and a constant velocity for all trials is therefore assumed.

The size of the time gap is drawn from a normal distribution with mean $\mu_{gap}$ and standard deviation $\sigma_{gap}$. Statistics for $\mu_{gap}$ and $\sigma_{gap}$ are obtained from results of the video graphic analysis by Feliciani et al. \cite{Feliciani2017}.

The ego vehicle's response to the pedestrian is simulated over Algorithm 1 for the parameters shown in Table \ref{tb:params}.

\begin{table}[h]
\begin{center}
\caption{Simulation and Controller Parameters}\label{tb:params}
\begin{tabular}{lccc}
Parameter & Symbol & Value & Units \\\hline\hline
Number of lanes & $n$ & 4 & - \\
Gap variance & $\sigma^2_{gap}$ & 2.5 & sec\\
Pedestrian velocity & $\dot{x}_p$ & 1.2 & $\mathrm{m/s}$ \\
Mean accepted gap & $\mu_{gap}$ & 4.0 & sec\\\hline
Safety offset       & $\Delta$ & 5.0 & m \\
Speed feedback gain & $k_s$ & 2.0 & 1 / sec\\
Brake time delay    & $t_\mathrm{delay}$ & 0 & sec \\ 
Speed limit & $v_\mathrm{speedlimit}$ & 4.5 & m/s \\
Comfort acceleration & $a_\mathrm{cmf}$ & 2 & $\mathrm{m/s^2}$ \\
Maximum time advantage & $t_\mathrm{max}$ & 4 & sec \\
Maximum acceleration & $a_\mathrm{max}$ & 9 & $\mathrm{m/s^2}$ \\\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Alternate Approach - POMDP}
The method proposed by Thornton \cite{Thornton2018} offers a point of comparison to the proposed algorithm, and models the unsignalized crosswalk problem as a Partially 
Observable Markov Decision Process (POMDP) \cite{Kochenderfer2015}. The relevant discretized state variable $x$ is given by

@@

Where $g_\mathrm{legality}$ and $g_\mathrm{safety}$ are terms encouraging the vehicle to stop when the pedestrian is in the crosswalk, $g_\mathrm{efficient}$ is a term encouraging high velocity when there is no pedestrian, and $g_\mathrm{smooth}$ encourages the vehicle to avoid abrupt changes in acceleration. Note that for our implementation of the method in \cite{Thornton2018}, we assumed $p_t$ was set to \textit{stopped}.

@@

Figure \ref{fig:velPlot1} shows simulation results for the case where the pedestrian enters the crosswalk on the same side of the road as the vehicle. The left column of the figure is for the case where the vehicle is on the right-most lane of a four way road (marked ``Lane A"), while the right column of the figure is for the case where the vehicle is in the second to right lane (Lane B). Every marker in the figure represents one of 1500 simulated vehicle crossings, 750 for each method.

@@

Figure \ref{fig:velPlot2} shows the same simulation results, but for the case where the pedestrian enters the crosswalk on the opposite side of the road as the vehicle. Again, the total number of pedestrian crossing trials conducted was 1500, with 750 trials for each method.

@@

The riskiest case occurs when the vehicle is in Lane A and the pedestrian crosses from the right side, accepting a gap between 1.25 and 1.75 seconds. Note that this is in general unlikely given gap acceptance statistics reported in the literature \cite{Rasouli}\cite{Feliciani2017}, but would represent a safety risk given the vehicle has limited time to react. However, since the algorithm recognizes that braking will not clear the crosswalk, the vehicle is able to speed up and maintain at least 2 meters of lateral distance from the pedestrian at all times. Note that a better way to confirm robustness of the controller to all reasonable pedestrian behaviors would be through a theoretical verification analysis - this will be the focus of future work.

In terms of the fifty state legality requirement, we conducted our simulation for the twenty states where the vehicle is required to yield, not stop, to a pedestrian in the same side of the crosswalk or approaching from the opposite side. For the US states where the vehicle is explicitly required to stop, the change to the controller is trivial - the controller may not stay in the driving mode if sufficient time advantage exits (see Algorithm 1).

@@

The requirement to minimize deviation from speed of traffic is validated in (c) and (d) of Figures ~\ref{fig:velPlot1} and ~\ref{fig:velPlot2}, which shows the average velocity of the vehicle for each simulated crossing. For high accepted gaps, the pedestrian simply waits for the ego vehicle to exit the crosswalk before beginning to walk, and the algorithm does not need to slow down from the speed of traffic. For low accepted gaps, the vehicle cannot stop, and must speed up or continue driving through the intersection, also keeping the vehicle speed high.

The most disruption to traffic flow occurs when the vehicle must slow down significantly or stop, which occurs for accepted gaps between 2-6 seconds. In this case, the average velocity increases as the accepted gap increases. Larger accepted gaps give the pedestrian more time to walk before the vehicle approaches, meaning they are closer to exiting the crosswalk when the vehicle must yield. This means the vehicle does not need to stop for as much time, resulting in higher average velocity.

Note that a special case occurs in Fig.~\ref{fig:velPlot2} where the vehicle is in Lane A and the pedestrian crosses from the opposite side of the road. Because the vehicle is at a far distance from the pedestrian, the time advantage remains sufficiently high for the vehicle to comfortably continue through the crosswalk without needing to slow down.

Table \ref{tb:speeds} shows a summary of the average speed compared to the baseline speed of 4.4 m/s for each of the four cases.

\begin{table}[h]
\begin{center}
\caption{Average Vehicle Speed}\label{tb:speeds}
\begin{tabular}{cccc}
Pedestrian Entry & Vehicle Lane & Hybrid Controller Average Speed (m/s) & POMDP Average Speed \\\hline\hline

Right Side & Right-Most & 2.90 & 66\% \\
Right Side & Second-to-Right & 2.93 & 67\% \\
Left Side  & Right-Most & 4.4 & 100\% \\
Left Side  & Second-to-Right & 2.80 & 64\% \\\hline

\end{tabular}
\end{center}
\end{table}

\subsubsection{Comparison with POMDP}
The policy from the POMDP displays interesting behavior at low and high pedestrian accepted gaps. At high accepted gaps, even though the pedestrian is waiting for the car to pass before crossing, the POMDP still slows the car down, resulting in a slight drop in average velocity. This is because the POMDP is accounting for the chance that the pedestrian might actually cross at a very low accepted gap. The closed loop POMDP policy does not appear to have knowledge that the vehicle can speed up at the last second. Meanwhile, the hybrid control policy has this case built in, and does not need to slow down if the pedestrian has not entered the crosswalk, resulting in an overall efficiency gain.

Another note is that the POMDP only models the pedestrian as being in the crosswalk or out of the crosswalk - this is necessary to limit the size of the discretized state space for the QMDP solver. This results in the POMDP having the same control policy across all four simulated cases. Because the proposed hybrid controller is continuous, there is no computational burden on accounting for the position and velocity of the pedestrian in the crosswalk. This allows for the controller to consider factors such as time advantage, and continue through the crosswalk in Fig.~\ref{fig:velPlot2}.

\subsection{Smoothness}
The requirement of keeping a low acceleration magntude where possible is validated through (e) and (f) of Figures~\ref{fig:velPlot1} and \ref{fig:velPlot2}, which plots the peak acceleration / deceleration magnitude for each simulated crossing. The vehicle is able to maintain a peak acceleration of 2 $\mathrm{m/s^2}$ for the vast majority of simulated trials, particularly when the pedestrian crosses from the opposing side of the road or when the vehicle is in the second to right line.

@@

\subsubsection{Comparison with POMDP}
Both the POMDP and the hybrid controller must decelerate at a high magnitude when a pedestrian accepts a gap between 1.75-2.5 seconds. One relative advantage of the POMDP is that it is able to decelerate at rates below $a_\mathrm{cmf}$ - at pedestrian accepted gaps of 5-6 seconds, the vehicle decelerates early and at a very low rate in order to avoid excessive jerk. However, by using the same policy for all four simulated scenarios and by not considering the ability to speed through the intersection, the POMDP also tends to decelerate when driving through the intersection would have sufficed.

@@

The experimental velocity and acceleration is plotted vs $d$ in 
Figure \ref{fig:expPlot}. In Trial 1, the virtual pedestrian accepts a gap of four seconds, giving the vehicle enough time to yield at a deceleration of 2 $\mathrm{m/s^2}$. Note that the low level acceleration controller exhibits a large time delay between commanded and actual acceleration. Because this is accounted for in Algorithm 1, the vehicle is able to brake early and the vehicle is stopped right at $d=0$ before the pedestrian exits the crosswalk and the vehicle speeds back up.

>>>>

As autonomous vehicles (AVs) inch closer to reality, a central requirement for acceptance will be earning the trust of humans in everyday driving situations. In particular, the interaction between AVs and pedestrians is of
high importance, as every human is a pedestrian at some point of the day. This paper considers the interaction of a pedestrian and an autonomous vehicle at a mid-block, unsignalized intersection where there is ambiguity
over when the pedestrian should cross and when and how the vehicle should yield. By modeling pedestrian behavior through the concept of gap acceptance, the authors show that a hybrid controller with just four distinct modes allows an autonomous vehicle to successfully interact with a pedestrian across a continuous spectrum of possible crosswalk entry behaviors. The controller is validated through extensive simulation and compared to an alternate POMDP solution and experimental results are provided on a research vehicle for a virtual pedestrian.

@@

While autonomous vehicles have the potential to save thousands of lives every year and create significant societal benefits \cite{Fagnant2015}, widespread adoption is unlikely until AVs gain the broad trust of society. Given that every human is a pedestrian at some point during the day, one of the central ways that autonomous vehicles will be evaluated is through their interactions with pedestrians. Interactions with pedestrians can be complex, even for experienced human drivers. From 2015-2016, pedestrian fatalaties increased by 9\% to 5987, representing the highest number since 1990, and also representing 16\% of all automotive fatalities \cite{HighwayTrafficSafetyAdministration2016}. As autonomous vehicles inch closer to widespread adoption, they must have a clear control strategy for pedestrian interaction that can handle a wide variety of pedestrian behaviors while maintaining a reasonable flow of traffic.

@@

The need to further understand pedestrian behavior for autonomous driving has created a signficant body of research focused on modeling pedestrian behavior given various sensor inputs. Keller et al. \cite{Keller2014} presented a study on pedestrian path prediction and action classification (e.g. crossing vs. waiting) using Gaussian process dynamical models and trajectory matching from optical data. Several other techniques for pedestrian trajectory prediction have been proposed, including LQR \cite{Batkovic}, set-based reachability analysis \cite{Koschi2018}, and Markov processes \cite{Karasev2016}. Another branch of literature focuses on modeling pedestrian crossing behavior in the form of video analysis. Schroeder and Rouphail \cite{Schroeder2011} explored factors associated with driver yielding behavior at unsignalized pedestrian crossings. Using logistic regression, the authors found that drivers are more likely to yield to assertive pedestrians who walk briskly in their approach to a crosswalk. Kadali and Perumal \cite{RaghuramKadali2012} studied the ``gap acceptance" behavior of pedestrians at mid-block crosswalks, and found that the gap accepted for crossing was explained by factors such as crossing direction, vehicle speed, and pedestrian age. Yannis et al. \cite{Yannis2013} and Sun et al.\cite{Sun2002} also found that gap acceptance was influenced by the size of the oncoming vehicle and the presence of other pedestrians.

@@

This paper aims to address the above issues by developing a hybrid control architecture that accounts for several distinct pedestrian modes of behavior at an unsignalized crosswalk. Simulations show that the proposed hybrid controller is able to handle a continuous spectrum of pedestrian \textit{gap acceptance} behavior, tolerating a range of highly conservative to highly aggressive pedestrians. Additionally, this paper provides a simulated comparison between the hybrid controller and the solution method proposed by Thornton \cite{Thornton2018}, citing advantages and disadvantages of each method. Finally, experimental results are shown on a real vehicle to demonstrate the feasibility of the proposed controller.

@@

\begin{figure}
\centering
\includegraphics[width=2.5in]{figures/example2.eps}
\caption{Top: Description of relevant problem variables. Bottom: Schematic of pedestrian approaching a stream of traffic at a mid-block intersection. Ego vehicle shown in green. We develop a hybrid control framework for the green vehicle, which considers the distance ($d$) to a stopping point $\Delta$ ahead of the crosswalk and the pedestrian's position in the crosswalk ($x_p$).}
\label{fig:schematic}
\end{figure}

@@

A diagram of the relevant state variables is shown in Fig.~\ref{fig:schematic}. We assume that knowledge of distance $d(t)$ to a fixed stopping point $\Delta$ 3-5 meters ahead of the crosswalk can be estimated by a perception system, and that the vehicle velocity $v(t)$ is known as well. We also assume the perception system is able to determine the position of the pedestrian $x_p$ within the crosswalk, and is able to determine a velocity estimate $\dot{x}_p$ for the pedestrian. Note that $x_p$ can be negative for the case where the pedestrian is on the sidewalk.

@@

\begin{equation}
a = k(d, v, x_p, \dot{x_p})
\end{equation}

@@

A difficulty with this control problem is that there are multiple stakeholders and therefore multiple competing objectives that must be considered. Relevant stakeholders for this problem are the autonomous vehicle occupants, the pedestrian(s), and the other vehicles in the traffic stream.

@@

The accepted gap is a measure of how much time there is before the vehicle would enter the crosswalk if it kept its current speed constant. When faced with a stream of traffic at an unsignalized crosswalk, a pedestrian inherently decides how much of a time gap to accept before crossing. The average gap acceptance is reported in the literature to be between 3-7 seconds, meaning pedestrians usually do not cross if the vehicle would enter the crosswalk in under three seconds \cite{DiPietroCharlesMandKing1970}, and are very likely to cross when they have more than seven seconds \cite{Schmidt2009}.

@@

The proposed hybrid control architecture is shown in Fig.~\ref{fig:hybridController}, and the algorithm for operation is shown in Algorithm \ref{alg:hybridController}. At its core, the control algorithm uses a feedback-feedforward methodology to compute desired acceleration commands, but computes these commands differently depending on which of four discrete modes the controller is in. The next section will describe the four modes.

@@

\begin{algorithm} % enter the algorithm environment
\caption{Calculate $a = k(d, v, x_p, \dot{x}_p)$} % give the algorithm a caption
\label{alg:hybridController} % and a label for \ref{} commands later in the document
\begin{algorithmic}
\State $state \leftarrow DRIVING$
\While {$\mathbf{true}$}
\If {$state = DRIVING$}
	\State $a \gets k_s(v-v_\mathrm{speed limit})$ 
	\If{$d > 0 \hspace{1mm} \mathbf{and} \hspace{1mm} \mathrm{inCrosswalk}(x_p)$}
		\If{$\mathrm{timeAdvantage}(x_p, \dot{x_p}, d, v) > \tau_\mathrm{max}$} 
		\State $state \leftarrow DRIVING$

@@

\subsection{Nominal Driving}
In the nominal state, the vehicle attempts to drive through the crosswalk at the speed limit $v_\mathrm{speedlimit}$. A proportional speed control is applied to keep the vehicle at the speed limit, achieving design constraint 2.a:

@@

The controller remains in the driving mode unless the pedestrian begins to enter the crosswalk before the vehicle has passed. We mathematically define ``beginning to enter" as the instant the pedestrian begins moving towards the crosswalk, even if the pedestrian is currently on the sidewalk:

@@

 Where $x_F$ is the end of the crosswalk area of interest. According to design constraint 1.b, the definition of $x_F$ will vary according to which US state the vehicle is in. For example, in California, $x_F$ would be at the end of the crosswalk, while in Louisiana, $x_F$ would just be the same half of the crosswalk as the vehicle \cite{Legislatures2018}. Once the pedestrian enters the crosswalk, the algorithm decides which mode to enter next.

@@

Frequently the vehicle will pass through the crosswalk well before the pedestrian will reach the vehicle's lane. Assuming the vehicle is operating in a US state where stopping is not explicitly required, design constraint 2.a suggests it is desirable for the vehicle to continue through the intersection.

This can be formalized by calculating the \textit{time advantage} $t_\mathrm{adv}$ \cite{Chen} as follows:

@@

Where $x_v$ is the $x$ position of the vehicle in the crosswalk. Algorithm 1 explicitly allows the vehicle to continue driving through the intersection as long as $t_\mathrm{adv}$ exceeds a specified threshold $\tau_\mathrm{max}$.

@@

In the unlikely event the pedestrian crosses when the time gap is low, the vehicle will need to prioritize design constraint 1.a and come to a stop at a deceleration higher than $-a_\mathrm{cmf}$. This occurs when the following conditions holds:

@@

As a final condition, consider the case where the pedestrian enters the crosswalk just as the vehicle is crossing $d = 0$. In this case, the vehicle has insufficient space to stop before the crosswalk and will risk being rear-ended or stopping in the crosswalk. It makes more sense in this condition for the vehicle to speed up and exit the crosswalk quickly. As a result, if $d < \frac{v}{2a_\mathrm{max}}$, the vehicle speeds up.

@@

Where $d_o$ and $v_o$ are the values of $d$ and $v$ when the controller first enters the yielding state. Inspection shows that the desired velocity is $v_o$ when $d = d_o$ and is 0 when $d = 0$. 
% (recall that we enter the yielding state when $d_o = \frac{\dot{d_o}^2}{2a_\mathrm{cmf}}$).

The controller exits the yielding mode and returns to the driving mode once the pedestrian clears the crosswalk.

@@

To validate the algorithm, both simulation and experimental studies were undertaken. The simulation results focused on testing Algorithm 1 against the specified design requirements over many pedestrian crossings.

The algorithm is validated in simulation via a built-from-scratch crosswalk simulation environment developed in Python (Fig.~\ref{fig:simFramework}). The simulation environment is built on the ROS robotics interface to enable easy porting of the control architecture to the experimental vehicle in Section \ref{sec:expres}, and for ease of comparison to the method provided by Thornton \cite{Thornton2018}.

\begin{figure}
\centering
\includegraphics[width=2.0in]{figures/simFramework.png}
\caption{Crosswalk simulation environment for algorithm validation. Ego vehicle is shown in black, leading vehicle in traffic stream shown in blue, and pedestrian shown in red.}
\label{fig:simFramework}
\end{figure}

To simulate random pedestrian crossing events, a pedestrian is spawned on the sidewalk near the entrance to the crosswalk. The pedestrian accepts a randomly sized time gap in the traffic stream to cross the road at constant speed $\dot{x}_p$. The size of the time gap is drawn from a normal distribution with mean $\mu_{gap}$ and standard deviation $\sigma_{gap}$. Statistics for $\mu_{gap}$ and $\sigma_{gap}$ are obtained from results of the video graphic analysis by Feliciani et al. \cite{Feliciani2017}. The ego vehicle's response to the pedestrian is simulated over Algorithm 1 for the parameters shown in Table \ref{tb:params}.

\begin{table}[h]
\begin{center}
\caption{Simulation and Controller Parameters}\label{tb:params}
\begin{tabular}{lccc}
Parameter & Symbol & Value & Units \\\hline\hline
Number of lanes & $n$ & 4 & - \\
Gap variance & $\sigma^2_{gap}$ & 2.5 & sec\\
Pedestrian velocity & $\dot{x}_p$ & 1.2 & $\mathrm{m/s}$ \\
Mean accepted gap & $\mu_{gap}$ & 4.0 & sec\\\hline
Safety offset    & $\Delta$ & 5.0 & m \\
Speed feedback gain & $k_s$ & 2.0 & 1 / sec\\
Brake time delay  & $t_\mathrm{delay}$ & 0 & sec \\ 
Speed limit & $v_\mathrm{speedlimit}$ & 4.5 & m/s \\
Comfort acceleration & $a_\mathrm{cmf}$ & 2 & $\mathrm{m/s^2}$ \\
Maximum time advantage & $\tau_\mathrm{max}$ & 4 & sec \\
Maximum acceleration & $a_\mathrm{max}$ & 9 & $\mathrm{m/s^2}$ \\\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Alternate Approach - POMDP}
The method proposed by Thornton \cite{Thornton2018} offers a comparison to the proposed algorithm and models the unsignalized crosswalk problem as a Partially 
Observable Markov Decision Process (POMDP) \cite{Kochenderfer2015}. The relevant discretized state variable $x$ is given by

@@

Where $g_\mathrm{legality}$ and $g_\mathrm{safety}$ are terms encouraging the vehicle to stop when the pedestrian is in the crosswalk, $g_\mathrm{efficient}$ is a term encouraging high velocity when there is no pedestrian, and $g_\mathrm{smooth}$ encourages the vehicle to avoid abrupt changes in acceleration. For our implementation of \cite{Thornton2018}, we assumed $p_t$ was set to \textit{stopped}.

@@

Figure \ref{fig:velPlot1} shows simulation results for the case where the pedestrian enters the crosswalk on the same side of the road as the vehicle. The vehicle is simulated in both Lane A and Lane B in Fig.~\ref{fig:simFramework}. Every marker in the figure represents one of 1500 simulated vehicle crossings, 750 for each method. Figure \ref{fig:velPlot2} shows the same simulation results, but for the case where the pedestrian enters the crosswalk on the opposite side of the road as the vehicle.

@@

@@

The riskiest case occurs when the vehicle is in Lane A and the pedestrian crosses from the right side, accepting a gap between 1.25 and 1.75 seconds. This is unlikely given gap acceptance statistics reported in the literature \cite{Rasouli}\cite{Feliciani2017}, but would represent a safety risk given the vehicle has limited time to react. However, since the algorithm recognizes that braking will not clear the crosswalk, the vehicle is able to speed up and maintain at least 2 meters of lateral distance from the pedestrian at all times. Future work will use formal verification methods in order to show safety for all pedestrian accepted gaps.

% In terms of the fifty state legality requirement, we conducted our simulation for the twenty states where the vehicle is required to yield, not stop, to a pedestrian in the same side of the crosswalk or approaching from the opposite side. For the US states where the vehicle is explicitly required to stop, the change to the controller is trivial - the controller may not stay in the driving mode if sufficient time advantage exits (see Algorithm 1).

@@

 Figures ~\ref{fig:velPlot1} and ~\ref{fig:velPlot2} show the average velocity of the vehicle for each simulated crossing in (c) and (d). For high accepted gaps, the pedestrian waits for the ego vehicle to exit the crosswalk before beginning to walk, and the algorithm does not need to slow down from the speed of traffic. For low accepted gaps, the vehicle cannot stop, and must speed up or continue driving through the intersection, also keeping the vehicle speed high.

The most disruption to traffic flow occurs when the vehicle must slow down significantly or stop, which occurs for accepted gaps between 2-6 seconds. A special case occurs in Fig.~\ref{fig:velPlot2} where the vehicle is in Lane A and the pedestrian crosses from the opposite side of the road. Because the vehicle is at a far distance from the pedestrian, the time advantage remains sufficiently high for the vehicle to comfortably continue through the crosswalk without needing to slow down.

\subsubsection{Comparison with POMDP}
The policy from the POMDP displays interesting behavior at low and high pedestrian accepted gaps. At high accepted gaps, even though the pedestrian is waiting for the car to pass before crossing, the POMDP still slows the car down. The POMDP, which uses a probabilistic model of pedestrian crossing, is accounting for the chance the pedestrian might cross at a very low accepted gap. The POMDP policy does not account for the possibility of the vehicle speeding up at the last second. The hybrid control policy has this case built in, and does not need to slow down if the pedestrian has not entered the crosswalk, resulting in an overall efficiency gain.

Also note the POMDP models the pedestrian as being in the crosswalk or out of the crosswalk - this is necessary to limit the size of the discretized state space for the QMDP solver. This results in the POMDP having the same control policy across all four simulated cases. Because the proposed hybrid controller is continuous, there is no computational burden on accounting for the position and velocity of the pedestrian in the crosswalk. This allows for the controller to consider factors such as time advantage, and continue through the crosswalk in many cases.

% Table \ref{tb:speeds} shows a summary of the average speed compared to the baseline speed of 4.4 m/s for each of the four cases.

% \begin{table}[h]
% \begin{center}
% \caption{Average Vehicle Speed}\label{tb:speeds}
% \begin{tabular}{cccc}
% Pedestrian Entry & Vehicle Lane & Hybrid Controller Average Speed (m/s) & POMDP Average Speed \\\hline\hline

% Right Side & Right-Most & 2.90 & 66\% \\
% Right Side & Second-to-Right & 2.93 & 67\% \\
% Left Side & Right-Most & 4.4 & 100\% \\
% Left Side & Second-to-Right & 2.80 & 64\% \\\hline

% \end{tabular}
% \end{center}
% \end{table}

\subsection{Smoothness}
Plots (e) and (f) of Figures~\ref{fig:velPlot1} and \ref{fig:velPlot2} show the peak acceleration / deceleration magnitude for each simulated crossing. The hybrid controller maintains a peak acceleration of 2 $\mathrm{m/s^2}$ for the vast majority of simulated trials, particularly when the pedestrian crosses from the opposing side of the road or when the vehicle is in Lane B.

@@

\subsubsection{Comparison with POMDP}
 One relative advantage of the POMDP is that it is able to decelerate at rates below $a_\mathrm{cmf}$. At pedestrian accepted gaps of 5-6 seconds, the vehicle decelerates early and at a very low rate in order to avoid excessive jerk. However, by using the same policy for all four simulated scenarios and by not considering the ability to drive through the intersection, the POMDP also tends to decelerate when driving through the intersection would have sufficed.

@@

The experimental velocity and acceleration is plotted vs $d$ in 
Figure \ref{fig:expPlot}. In Trial 1, the pedestrian accepts a gap of four seconds, giving the vehicle enough time to yield at a deceleration of 2 $\mathrm{m/s^2}$. Note that the low level longitudinal controller exhibits a large time delay between commanded and actual acceleration. Because this is accounted for in Algorithm 1, the vehicle is able to brake early and the vehicle is stopped right at $d=0$ before the pedestrian exits the crosswalk and the vehicle speeds back up.
